Classless Java: Tuning Java Interfaces

### 2015-11-24 paper/paper.tex
Ok, pdflatex is now working fine. I submit a couple of typo fix to introduction/abstract and some new text.
Proof in progress.
proof in appendix.
Minor changes in Introduction
Formalization partly revised
changes

\newcommand\mixin{\textbf{@Mixin} }
\usepackage{xspace}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Classless Java}
\titlerunning{Classless Java}

@@

\Copyright{John Q. Open and Joan R. Access}
% mandatory, please use full first names. LIPIcs license is "CC-BY";
% http://creativecommons.org/licenses/by/3.0/

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\serieslogo{}%please provide filename (without suffix)
\volumeinfo%(easychair interface)
  {Billy Editor and Bill Editors}% editors
  {2}% number of editors: 1, 2, ....
  {Conference title on which this volume is based on}% event
  {1}% volume
  {1}% issue
  {1}% starting page number
\EventShortName{}
\DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This paper shows that, with some minor syntactic sugar, default
methods and interface inheritance are in fact be very expressive. Our
proposed syntax sugar, called \emph{object interfaces}, enables
powerful object-oriented idioms, using multiple inheritance, to be
expressed directly in Java. Object interfaces avoid some syntactic
boilerplate caused by Java being originally designed to work mainly
with classes. Moreover, object interfaces come with their own object
instantiation mechanism, providing an alternative to class
constructors. As a result many Java programs can be built without
defining a single class! An implementation of object interfaces using 
Java annotations and a formalization of the static and dynamic
semantics are presented. Moreover the usefulness of object
interfaces is illustrated through various examples and case studies.
\bruno{a first draft of the abstract. still missing ``lightweight
  language extensions''.}

@@

@@

\appendix

>>>>

\newcommand\mixin{\textbf{@Mixin} }
\renewcommand{\paragraph}[1]{\vspace{5pt}\noindent{\bf #1}}
\usepackage{xspace}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Classless Java: Tuning Java Interfaces}
\titlerunning{Classless Java}

@@

\begin{comment}
\Copyright{John Q. Open and Joan R. Access}
% mandatory, please use full first names. LIPIcs license is "CC-BY";
% http://creativecommons.org/licenses/by/3.0/

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\serieslogo{}%please provide filename (without suffix)
\volumeinfo%(easychair interface)
  {Billy Editor and Bill Editors}% editors
  {2}% number of editors: 1, 2, ....
  {Conference title on which this volume is based on}% event
  {1}% volume
  {1}% issue
  {1}% starting page number
\EventShortName{}
\DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{comment}

@@

This paper shows that, with a simple language feature, default methods
and interface inheritance are in fact very expressive. Our proposed
language feature, called \emph{object interfaces}, enables powerful
object-oriented idioms, using multiple inheritance, to be expressed
conveniently in Java. Object interfaces refine conventional Java
interfaces in three different ways. Firstly, object interfaces have
their own object instantiation mechanism, providing an alternative to
class constructors. Secondly, object interfaces support \emph{abstract
  state operations}, providing a way to use multiple inheritance with
state in Java. Finally, object interfaces allow type refinements that
are often tricky to model in conventional class-based
approaches. Interestingly, object interfaces do not require changes to
the runtime, and they also do not introduce any new syntax (apart from
a small annotation): all three features are achieved by reinterpreting
existing Java syntax, and are translated into regular Java code
without loss of type-safety. Since no new syntax is introduced, it
would be incorrect to call object interfaces a language extension or
syntactic sugar. So we use the term \emph{language tuning} to
characterize this kind of language feature. An implementation of
object interfaces using Java annotations and a formalization of the
static and dynamic semantics are presented. Moreover the usefulness of
object interfaces is illustrated through various examples.

@@

\input{sections/Translation.tex}

@@

\appendix
\input{sections/Appendix.tex}

### 2015-11-24 paper/sections/Introduction.tex
changes

Still, the design has been criticised for being too conservative and 
not very useful in its current form to model traits or multiple 
inheritance~\cite{}. 
Indeed, our own personal experience of combining default methods 
and multiple interface inheritance in Java to achieve multiple implementation 
inheritance is that many workarounds and boilerplate code are needed. 
In particular, we encountered difficulties because:

@@

\item {\em Interfaces do not have state.} This creates a tension between 
 using multiple inheritance and having state. Using setter and
  getter methods is a way out of this tension, but this workaround
  requires tedious boilerplate classes that latter implements those
  methods.

@@

\noindent Clearly, a way around those difficulties would be to try to
change Java and just remove these limitations. Scala's own
notion of traits, for example, allows state in traits. Of course adding
state (and other features) to interfaces would complicate the language
and require changes in the compiler, and this would go beyond the
goals of Java 8 development team.

This paper takes a different approach. Rather than trying to get
around the difficulties by changing the language in fundamental ways,
we show that, with some minor syntactic sugar, default methods and
interface inheritance are in fact be very expressive. Our proposed
syntax sugar enables powerful object-oriented idioms, using multiple
inheritance, to be expressed directly without modifications to the
current Java language. We call our sugar \emph{object interfaces},
because such interfaces can be instantiated directly, without the need
for an explict class definition. Object interfaces come with their own
object instantiation mechanism, providing an alternative to class
constructors. Moreover, object interfaces avoid some syntactic
boilerplate caused by Java being originally designed to work mainly
with classes. This includes support for various common utility methods
(such as getters and setters, or clone methods), even in the presence
of subtyping. As a result, with object interfaces, many Java programs
can be built without using a single class!  \bruno{integrate more of
  Marco's text.}

To formalize object interfaces, we propose Classless Java (CJ)\marco{I propose to call it directly ClassLess Java, the  Featherweight part is in being classless.}: 
a FeatherweightJava-style calculus, which captures the essence 
of interfaces with default methods. The semantics of object interfaces 
is given as a type-directed translation from CJ to itself. 
In the resulting CJ code all object interfaces are translated into 
regular CJ (and Java) interfaces with default methods. The translation 
is proved to be type-preserving, ensuring that the translation does 
not introduce type-errors. One interesting aspect of the theorem is that, 
if we ignore the object annotations, the program still type-checks 
under the regular typing rules of CJ (and Java).\bruno{double-check!}
\marco{ I do not understand what you mean. 1) annotations are part of java and CJ, so you do not need to remove them 2) programs well typed without expansion are well typed with expansion, here you seams to state the opposite}
CJ's usefulness goes beyond serving as a calculus to formalize object
interfaces. During the development process of CJ, we encountered an 
incoherent behaviour in Java's implementation of default methods that 
could be classified as a bug. \bruno{Should we report bug to the Java
  developers mailing list?}
\marco{we need to put the exact javac version}
We also present a prototype implementation of object interfaces using
Java annotations. One benefit of an annotation-based implementation 
is that existing Java tools (such as IDE's) work with out-of-the-box
with our implementation. As a result we could experiment object
interfaces with several interesting Java programs, and conduct various
case studies. 
\marco{ I removed your text here and I add after to try talking about tuning}
%I think that the qualifier object goes AGAINST the idea of tuning.
%\bruno{mention the following or not? maybe good if we
%  had something about tuning.} Of course, another implementation option would be to
%extend Java, with a single qualifier keyword {\bf object}, so that
%object interfaces would be an intrinsic part of the language. We leave
%this for future work.

To evaluate the usefulness of object interfaces, we illustrate 3
applications and case studies. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application is to show how embedded DSLs using fluent interfaces~\cite{} 
can be easily defined using object interfaces. Finally, the last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, is the fact that all applications are implemented 
without defining a single class.

Moreover, we believe that our approach is an example of a more general idea, that we call \emph{Language Tuning},
a concept that sits in the middle between a lightweight language extension and a glorified library.
Thanks to Java dynamic class loading and annotation processing,  annotations can trigger type-driven transformations on classes during either loading or compilation.
There are many examples of such Language Tuning libraries,~\cite[lomboc,anotherImSearching].
Language tuning can offers many features usually implemented by a real language extension, but they do not modify the syntax; pre-existing tools can work transparently on the tuned language.

@@

\item {\bf Object Interfaces:} A simple feature that allows 
various powerful programming idioms to be expressed conveniently in
Java. Object interfaces are pure sugar: they are desugared into standard Java 
code, without loss of modularity or type-safety.

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the desugaring of object interfaces. 
We prove ...

\item {\bf A Bug in the Implementation of Java 8:}\bruno{state this as
  a contrib?}\marco{yes, especially if we "suggest solutions"}

@@

\item{\bf Language Tuning} We identify this concept and we describe Project Lomboc and object interfaces as examples of Language Tuning.
\end{itemize}

>>>>

Still, the design is quite conservative and appears to be quite limited
 in its current form to model advanced forms of multiple inheritance.
Indeed, our own personal experience of combining default methods 
and multiple interface inheritance in Java to achieve multiple implementation 
inheritance is that many workarounds and boilerplate code are needed. 
In particular, we encountered difficulties because:

@@

\item {\em Interfaces do not have state.} This creates a tension between 
 using multiple inheritance and having state. Using setter and
  getter methods is a way out of this tension, but this workaround
  requires tedious boilerplate classes that latter implement those
  methods.

@@

\noindent Clearly, a way around those difficulties would be to
change Java and just remove these limitations. Scala's own
notion of traits, for example, allows state in traits. Of course adding
state (and other features) to interfaces would complicate the language
and require changes in the compiler, and this would go beyond the
goals of Java 8 development team.

This paper takes a different approach. Rather than trying to get
around the difficulties by changing the language in fundamental ways,
we show that, with a simple language feature, default methods and
interface inheritance are in fact be very expressive. Our proposed
language feature enables powerful object-oriented idioms, using
multiple inheritance. We call the language feature \emph{object
  interfaces}, because such interfaces can be instantiated directly,
without the need for an explict class definition. Moreover, object
interfaces support \emph{abstract state operations}, providing a way
to use multiple inheritance with state in Java. The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately. With object interfaces, many Java programs can be built
without using a single class!

Object interfaces do not require changes to the
runtime, and they also do not introduce any new syntax (apart from a
small annotation). All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full-benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations and AST
rewriting, allowing existing Java tools (such as IDE's) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a type-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-preserving, ensuring that the translation does not
introduce type-errors. CJ's usefulness goes beyond serving as a
calculus to formalize object interfaces. During the development
process of CJ, we encountered a bug in the implementation of default
methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
bug, ECJ behaves differently from both our formalization and Oracle's 
Java compiler.

To evaluate the usefulness of object interfaces, we illustrate 3
applications and case studies. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application is to show how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. Finally, the last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, is the fact that all applications are implemented 
without defining a single class.

@@

\item {\bf Object Interfaces:} A simple feature that allows various
  powerful multiple-inheritanc eprogramming idioms to be expressed
  conveniently in Java.

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove a type-preservation theorem, and we present a Java program 
that reveals a bug in the ECJ implementation of default methods.

@@

\item{\bf Language Tuning} We identify this concept and we describe
object interfaces as examples of Language Tuning. We also discuss 
how other existing approaches, such as the annotations in project 
Lombok~\cite{lombok}, can be viewed as language tuning.

### 2015-11-24 paper/sections/Overview.tex
changes

\marco{fig1 is now better. Where we cite it in the text? can we make it so that is more "linear"? point2d <- point3D,   animal <- (horse over bird ) <- pegasus}
To propose a standard example, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ as a kind of \Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

>>>>

To propose a standard example, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ as a kind of \Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}
